The Stakes of Scrying: \\ \lstinline[style=inlinecode]{.^} dotket and the Seer Proposal

### 2024-03-08 mss.tex
Post HTML

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              Manuscript submitted for review.\\
              % {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
              ~ \\
              Address author correspondence to \patp{rovnys-ricfer}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{The Stakes of Scrying:  \lstinline[style=inlinecode]{.^} dotket and the Seer Proposal}
\fancyfoot[LE,RO]{\thepage}

@@

%%MANUSCRIPT
\title{The Stakes of Scrying:  \lstinline[style=inlinecode]{.^} dotket and the Seer Proposal}
\author{\authorname~\authorpatp \\ \affiliation}
\date{}

@@

\begin{abstract}
Nock 12 and the \dotket~rune break the referential transparency of Nock, with consequences for the purity of Urbit's functional programming model.  We consider the implications of this, and propose a move-based replacement for the \dotket~rune that would restore referential transparency to Nock.
\end{abstract}

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{1}

@@

While Nock formally has twelve opcodes, numbered 0--11, it is convenient to expose in userspace a fake Nock 12 opcode to expose the scry namespace to userspace developers.  This opcode is trivially invoked by using the Hoon \dotket~rune.  However, Nock 12 breaks the referential transparency of Nock, and this has consequences for the purity of Urbit's functional programming model.

\section[How dotket Works]{How \dotket~Works}

The Nock 12 opcode and the \dotket~rune are used by userspace code to ask the kernel for data in the scry namespace.\footnote{The scry namespace, in brief, is a collection of immutable values “bound” by convention at fixed endpoints.  The scry namespace is defined by a unique path for each resource.  A representative scry path would have the entries \lstinline[style=inlinecode]{/host/rift/life/vane/request-type/revision-number/desk/file-path/mark}.}  When Urbit runs userspace code, that code is not raw Nock.  It is instead called Mock, which is Nock with two additions:   deterministic errors are caught and returned as values to the kernel; and there is an extra Nock opcode, opcode 12.  A simple Nock 12 formula looks like \lstinline[style=inlinecode]{[12 [1 /foo/bar/baz]]}, where \lstinline[style=inlinecode]{/foo/bar/baz} is a path literal \lstinline[style=inlinecode]{[%foo %bar %baz ~]}, representing a path in the scry namespace.  The result of running this opcode will be the result of performing the scry request, i.e. dereferencing the path.

@@

As mentioned, the \lstinline[style=inlinecode]{[subject formula]} pair is not the only input to \lstinline[style=inlinecode]{++mink}.  There is also an argument of the form \lstinline[style=inlinecode]{scry=$-(^ (unit (unit)))}, called the scry handler gate.  When \lstinline[style=inlinecode]{++mink} encounters a Nock 12 opcode while evaluating the \lstinline[style=inlinecode]{[subject formula]} expression, it evaluates the sub-formula to obtain a path to look up in the scry namespace.  Once it has this path, it calls the scry handler gate on the path, which returns one of:

@@

The Nock 12 opcode is problematic because it renders the result of the computation dependent on external data which are not in the \lstinline[style=inlinecode]{[subject formula]} that is being evaluated.  The result is no longer a pure function of the \lstinline[style=inlinecode]{subject} and \lstinline[style=inlinecode]{formula}:  depending on what the scry handler gate is, the same \lstinline[style=inlinecode]{[subject formula]} could yield different results.  Consider \lstinline[style=inlinecode]{[subject formula]} of \lstinline[style=inlinecode]{[0 [12 1 /foo/bar]]}.  This returns the result of looking up the \lstinline[style=inlinecode]{/foo/bar} scry request path from the scry gate. Userspace code (which is untrusted) could call \lstinline[style=inlinecode]{(mink [[0 12 1 /foo/bar]] |=(^ [~ ~ 33]))}, acquire the result \lstinline[style=inlinecode]{33}, then call \lstinline[style=inlinecode]{(mink [[0 12 1 /foo/bar]] |=(^ [~ ~ 44]))} to yield \lstinline[style=inlinecode]{44}.  The \lstinline[style=inlinecode]{subject}s and \lstinline[style=inlinecode]{formula}s were identical, but because they were evaluated with different scry gates, they yielded different results.

@@

The Arvo kernel's scry handler gate, as described in the original Moron Lab blog post about Urbit \citep{Yarvin2010}, is a ``static functional namespace'', so even though that gate changes frequently, it maintains the invariant that every scry path's binding to a value is immutable.\footnote{Up to some known flaws that are addressed by UIP-0116: Arvo Ticks \citep{Monk2023, Monk2023a, Wilson2024}.}

@@

The current reference implementation of Seer is available at \citet{Levan2023}.  A more complete example is available at \citet{Levan2023a}.

@@

\printbibliography
\end{document}

>>>>

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:1 (2024):  35–46. \\
              ~ \\
              Address author correspondence to \patp{rovnys-ricfer}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{\lstinline[style=inlinecode]{.^} dotket and the Seer Proposal}
\fancyfoot[LE,RO]{\thepage}

@@

%%MANUSCRIPT
\title{The Stakes of Scrying:  \\ \lstinline[style=inlinecode]{.^} dotket and the Seer Proposal}
\author{\authorname~\authorpatp \\ \affiliation}
\date{}

@@

\begin{abstract}
\noindent
The expedient of Nock 12 and the \dotket~rune break the referential transparency of Nock, with consequences for the purity of Urbit's functional programming model.  We consider the implications of this and propose a move\-based replacement for the \dotket~rune that restores referential transparency.
\end{abstract}

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{35}

@@

While Nock formally has twelve opcodes, numbered 0--11, it is convenient to expose in userspace a fake Nock 12 opcode to expose the scry namespace to userspace developers without an asynchronous retrievel step.  This opcode is trivially invoked by using the Hoon \dotket~rune.\footnote{The idea of using a pattern like scrying to introduce a value into a referentially transparent namespace predates Urbit itself; cf. \citet{Yarvin2006}.}  However, Nock 12 breaks the referential transparency of Nock, and this has consequences for the purity of Urbit's functional programming model.

\section{How \dotket~Works}

The Nock 12 opcode and the \dotket~rune are used by user-\newline{}space code to ask the kernel for data in the scry namespace.\footnote{The scry namespace, in brief, is a collection of immutable values “bound” by convention at fixed endpoints.  The scry namespace is defined by a unique path for each resource.  A representative scry path would have the entries \texttt{/host/rift/life/vane/request-type/revision-number/desk\newline{}/file-path/mark}.}  When Urbit runs userspace code, that code is not raw Nock.  It is instead called Mock, which is Nock with two additions:   deterministic errors are caught and returned as values to the kernel; and there is an extra Nock opcode, opcode 12.  A simple Nock 12 formula looks like \lstinline[style=inlinecode]{[12 [1 /foo/bar/baz]]}, where \lstinline[style=inlinecode]{/foo/bar/baz} is a path literal \lstinline[style=inlinecode]{[%foo %bar %baz ~]}, representing a path in the scry namespace.  The result of running this opcode will be the result of performing the scry request, i.e. dereferencing the path.

@@

As mentioned, the \lstinline[style=inlinecode]{[subject formula]} pair is not the only input to \lstinline[style=inlinecode]{++mink}.  There is also an argument of the form \lstinline[style=inlinecode]{scry= $-(^ (unit (unit)))}, called the scry handler gate.  When \lstinline[style=inlinecode]{++mink} encounters a Nock 12 opcode while evaluating an expression \lstinline[style=inlinecode]{[subject formula]}, it evaluates the sub-formula to obtain a path to look up in the scry namespace.  Once it has this path, it calls the scry handler gate on the path, which returns one of:

@@

The Nock 12 opcode is problematic because it renders the result of the computation dependent on external data which are not in the \lstinline[style=inlinecode]{[subject formula]} that is being evaluated.  The result is no longer a pure function of the \lstinline[style=inlinecode]{subject} and \lstinline[style=inlinecode]{formula}:  depending on what the scry handler gate is, the same \lstinline[style=inlinecode]{[subject formula]} could yield different results.  Consider \lstinline[style=inlinecode]{[subject formula]} of \lstinline[style=inlinecode]{[0 [12 1 /foo/bar]]}.  This returns the result of looking up the \lstinline[style=inlinecode]{/foo/bar} scry request path from the scry gate. Userspace code (which is untrusted) could call \lstinline[style=inlinecode]{(mink [[0 12 1 /foo/bar]] |=(^ [~ ~ 33]))}, acquire a result \lstinline[style=inlinecode]{33}, then call \lstinline[style=inlinecode]{(mink [[0 12 1 /foo/bar]] |=(^ [~ ~ 44]))} to yield \lstinline[style=inlinecode]{44}.  The \lstinline[style=inlinecode]{subject}s and \lstinline[style=inlinecode]{formula}s were identical, but because they were evaluated with different scry gates, they yielded different results.

@@

\sloppy
The Arvo kernel's scry handler gate, as described in the original Moron Lab blog post about Urbit \citep{Yarvin2010}, is a ``static functional namespace''; even though that gate changes frequently, it maintains the invariant that every scry path's binding to a value is immutable.\footnote{Up to some known concerns that are addressed by UIP-0116: Arvo Ticks \citetpr{UIP-0116}.}

@@

The current reference implementation of Seer is available at \citetpr{Levan2023}.  A more complete example is available at \citealt{Levan2023a}.

@@

\selectlanguage{USenglish}
\printbibliography
\end{document}
